---
title: "Generation of simple time series: AR(1) process with white noise, incl. special case random walk"
author: Henk van Elst
date: Do, 27.5.2021
output:
  html_notebook:
    number_sections: true
    toc: true
    theme: united
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--#########################################################################-->
# Introduction
<!--#########################################################################-->
For a general introduction to time series analysis with R, see, e.g., URL: https://rc2e.com/timeseriesanalysis.

R Version `r getRversion()`.

<!--#########################################################################-->
# Load packages
<!--#########################################################################-->

```{r libraries}
library(tidyverse)
library(xts)
library(forecast)
library(urca)
```

<!--#########################################################################-->
# Generate synthetic time series
<!--#########################################################################-->
Setup a routine to generate time series data from an autoregressive process of
order one, AR(1), with either Gaußian or Gossetian white noise..

## Specify parameter values

```{r parameters}
n <- 250
alpha <- 0.75
sigma <- 2.0
df <- 3
type <- "gosset" # "gauss"
```

## Set time index

```{r init}
time <- Sys.Date() + 0:(n - 1)
```

## Divise AR(1) process with white noise

White noise can either be Gaußian (type: "gauss"; default) or Gossetian (type: 
"gosset").

```{r aroneprocess}
ar_one_process <-
  function(n,
           y0 = 0,
           alpha = 1,
           sigma = 1,
           df = 3,
           type = "gauss") {
    
    y <- vector(mode = "numeric", length = n)
    y[1] <- y0
    
    if (type == "gauss") {
      # Gaussian white noise
      w <- rnorm(n = n, mean = 0, sd = sigma)
    } else if (type == "gosset") {
      # Gossetian white noise
      w <- rt(n = n , df = df)
    }
    
    for (i in 2:n) {
      y[i] <- alpha * y[i - 1] + w[i]
    }
    
    return(y)
    
  }
```

## Generate time series data

```{r timeseries}
data <-
  tibble::tibble(
    y1 = ar_one_process(
      n = n,
      y0 = 0,
      alpha = 0.0,
      sigma = sigma,
      df = df,
      type = type
    ),
    y2 = ar_one_process(
      n = n,
      y0 = 0,
      alpha = -alpha,
      sigma = sigma,
      df = df,
      type = type
    ),
    y3 = ar_one_process(
      n = n,
      y0 = 0,
      alpha = 1.0,
      sigma = sigma,
      df = df,
      type = type
    ),
    y4 = ar_one_process(
      n = n,
      y0 = 0,
      alpha = alpha,
      sigma = sigma,
      df = df,
      type = type
    )
  )
```

## Collect data in xts object

```{r xtsobject}
y_xts <-
  xts::xts(
    x = data,
    order.by = time,
    frequency = 1,
    unique = TRUE
  )
```

<!--#########################################################################-->
# Lines representing quantiles of N(0, sigma^2), scaled by sqrt(t) according to random walk
<!--#########################################################################-->

```{r quantilelines}
quantileLineNorm <-
  function(x,
           sign = 1,
           prob = 0.89,
           sigma = 1.0) {
    t1 <- (x - Sys.Date())
    t2 <- as.numeric(t1)
    out <- sign * qnorm(p = prob, mean = 0, sd = sigma) * sqrt(t2)
    return(out)
  }
```

<!--#########################################################################-->
# Visualisation
<!--#########################################################################-->

```{r plot1}
plot.xts(
  x = y_xts ,
  main = "Simple time series models" ,
  xlab = "time" ,
  ylab = "displacement"
)
```

With ggplot2
URL: https://stackoverflow.com/questions/38597464/multiple-time-series-in-ggplot2-from-xts-object

```{r plot2}
mseries1 <-
  cbind(y_xts$y1 ,
        y_xts$y2 ,
        y_xts$y3 ,
        y_xts$y4)
names(mseries1) <-
  c("white noise" ,
    "negative alpha AR_1" ,
    "random walk" ,
    "positive alpha AR_1")
index(mseries1) <-
  as.Date(index(mseries1))

broom::tidy(x = mseries1) %>%
  ggplot(data = .,
         mapping = aes(x = index, y = value, colour = series)) +
  geom_line(size = 1.0) +
  labs(title = "Simple time series models") +
  xlab(label = "time [days]") +
  ylab(label = "displacement [1]") +
  stat_function(
    fun = quantileLineNorm,
    args = list(sign = 1,
                prob = 0.89,
                sigma = sigma),
    linetype = "dashed",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(
      sign = -1,
      prob = 0.89,
      sigma = sigma
    ),
    linetype = "dashed",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(sign = 1,
                prob = 0.97,
                sigma = sigma),
    linetype = "dotdash",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(
      sign = -1,
      prob = 0.97,
      sigma = sigma
    ),
    linetype = "dotdash",
    colour = "red"
  ) +
  theme_bw()
```

<!--#########################################################################-->
# Autocorrelation and partial autocorrelation
<!--#########################################################################-->

## White noise

```{r acfwn}
forecast::tsdisplay(
  x = y_xts$y1,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "White noise"
)
```

## AR(1): negative coeff, 0 < absVal < 1

```{r acfar1neg}
forecast::tsdisplay(
  x = y_xts$y2,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "AR(1), negative coeff, 0 < absVal < 1"
)
```


## AR(1): random walk

```{r acfrw}
forecast::tsdisplay(
  x = y_xts$y3,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "AR(1), random walk"
)
```


## AR(1): positive coeff, 0 < absVal < 1

```{r acfar1pos}
forecast::tsdisplay(
  x = y_xts$y4,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "AR(1), positive coeff, 0 < absVal < 1"
)
```

<!--#########################################################################-->
# Stationarity tests
<!--#########################################################################-->

## Augmented Dickey-Fuller test

```{r adf}
coredata(y_xts) %>%
  apply(
    X = .,
    MARGIN = 2,
    FUN = ur.df,
    type = "none",
    #lags = 32 ,
    selectlags = "AIC"
  ) %>%
  map(function(urca.element) {
    return(
      tibble(
        critical_values = as.data.frame(urca.element@cval) %>%
          filter(row_number() == 1),
        test_statistic = urca.element@teststat
      )
    )
  }) %>%
  map(function(results_table) {
    results_table %>%
      transmute(
        .data = .,
        `0.01` = .$critical_values$`1pct`,
        `0.05` = .$critical_values$`5pct`,
        `0.10` = .$critical_values$`10pct`,
        teststat = .$test_statistic
      )
  }) %>%
  bind_rows(.id = "time_series")
```

## Kwiatkowski–Phillips–Schmidt–Shin-test

```{r kpss}
coredata(y_xts) %>%
  apply(
    X = .,
    MARGIN = 2,
    FUN = ur.kpss,
    type = "mu" ,
    lags = "long"
  ) %>%
  map(function(urca.element) {
    return(
      tibble(
        critical_values = as.data.frame(urca.element@cval) %>%
          filter(row_number() == 1),
        test_statistic = urca.element@teststat
      )
    )
  }) %>%
  map(function(results_table) {
    results_table %>%
      transmute(
        .data = .,
        `0.01` = .$critical_values$`1pct`,
        `0.05` = .$critical_values$`5pct`,
        `0.10` = .$critical_values$`10pct`,
        teststat = .$test_statistic
      )
  }) %>%
  bind_rows(.id = "time_series")
```

<!--#########################################################################-->
# Generate synthetic lag-1-difference time series
<!--#########################################################################-->

```{r timeseriesdiff}
y_diff_xts <- diff(x = y_xts, lag = 1) %>%
  na.omit(object = .)
```

<!--#########################################################################-->
# Visualisation
<!--#########################################################################-->

```{r plot3}
plot.xts(
  x = y_diff_xts ,
  main = "Simple lag-1-difference time series models" ,
  xlab = "time" ,
  ylab = "lag-1-difference displacement"
)
```

ggplot2

URL: https://stackoverflow.com/questions/38597464/multiple-time-series-in-ggplot2-from-xts-object

```{r plot4}
mseries2 <-
  cbind(y_diff_xts$y1 ,
        y_diff_xts$y2 ,
        y_diff_xts$y3 ,
        y_diff_xts$y4)
names(mseries2) <-
  c(
    "white noise lag 1" ,
    "negative alpha AR_1 lag 1" ,
    "random walk lag 1" ,
    "positive alpha AR_1 lag 1"
  )
index(mseries2) <-
  as.Date(index(mseries2))

broom::tidy(x = mseries2) %>%
  ggplot(data = .,
         mapping = aes(x = index, y = value, colour = series)) +
  geom_line(size = 1.0) +
  labs(title = "Simple lag-1-difference time series models") +
  xlab(label = "time [days]") +
  ylab(label = "lag-1-difference displacement [1]") +
  stat_function(
    fun = quantileLineNorm,
    args = list(sign = 1,
                prob = 0.89,
                sigma = sigma),
    linetype = "dashed",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(
      sign = -1,
      prob = 0.89,
      sigma = sigma
    ),
    linetype = "dashed",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(sign = 1,
                prob = 0.97,
                sigma = sigma),
    linetype = "dotdash",
    colour = "red"
  ) +
  stat_function(
    fun = quantileLineNorm,
    args = list(
      sign = -1,
      prob = 0.97,
      sigma = sigma
    ),
    linetype = "dotdash",
    colour = "red"
  ) +
  theme_bw()
```

<!--#########################################################################-->
# Autocorrelation and partial autocorrelation
<!--#########################################################################-->

## White noise

```{r acfwndiff}
forecast::tsdisplay(
  x = y_diff_xts$y1,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "Lag-1-difference white noise"
)
```

## AR(1): negative coeff, 0 < absVal < 1

```{r acfa1negdiff}
forecast::tsdisplay(
  x = y_diff_xts$y2,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "Lag-1-difference AR(1), negative coeff, 0 < absVal < 1"
)
```

## AR(1): random walk

```{r acfrwdiff}
forecast::tsdisplay(
  x = y_diff_xts$y3,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "Lag-1-difference AR(1), random walk"
)
```

## AR(1): positive coeff, 0 < absVal < 1

```{r acfa1posdiff}
forecast::tsdisplay(
  x = y_diff_xts$y4,
  plot.type = "partial",
  points = TRUE,
  ci.type = "white",
  na.action = na.pass,
  main = "Lag-1-difference AR(1), positive coeff, 0 < absVal < 1"
)
```

<!--#########################################################################-->
# Stationarity tests
<!--#########################################################################-->

## Augmented Dickey-Fuller test

```{r adf2}
coredata(y_diff_xts) %>%
  apply(
    X = .,
    MARGIN = 2,
    FUN = ur.df,
    type = "none",
    #lags = 32 ,
    selectlags = "AIC"
  ) %>%
  map(function(urca.element) {
    return(
      tibble(
        critical_values = as.data.frame(urca.element@cval) %>%
          filter(row_number() == 1),
        test_statistic = urca.element@teststat
      )
    )
  }) %>%
  map(function(results_table) {
    results_table %>%
      transmute(
        .data = .,
        `0.01` = .$critical_values$`1pct`,
        `0.05` = .$critical_values$`5pct`,
        `0.10` = .$critical_values$`10pct`,
        teststat = .$test_statistic
      )
  }) %>%
  bind_rows(.id = "time_series")
```

## Kwiatkowski–Phillips–Schmidt–Shin-test

```{r kpss2}
coredata(y_diff_xts) %>%
  apply(
    X = .,
    MARGIN = 2,
    FUN = ur.kpss,
    type = "mu" ,
    lags = "long"
  ) %>%
  map(function(urca.element) {
    return(
      tibble(
        critical_values = as.data.frame(urca.element@cval) %>%
          filter(row_number() == 1),
        test_statistic = urca.element@teststat
      )
    )
  }) %>%
  map(function(results_table) {
    results_table %>%
      transmute(
        .data = .,
        `0.01` = .$critical_values$`1pct`,
        `0.05` = .$critical_values$`5pct`,
        `0.10` = .$critical_values$`10pct`,
        teststat = .$test_statistic
      )
  }) %>%
  bind_rows(.id = "time_series")
```

<!--#########################################################################-->
# Information on session
<!--#########################################################################-->

```{r session, echo = FALSE}
sessionInfo()
```

<!--#########################################################################-->
<!--#########################################################################-->
